---
title: "Game Metrics: Decoding Premier League Outcomes Through Visual Analysis and Data Exploration"
subtitle: "INFO 526 - Project 1"
author: "Group 4 WackyWednesday"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false
#| output: false
#| echo: false
if(!require(pacman))
  install.packages("pacman")

pacman::p_load(
  tidyverse,
  here,
  ggthemes,
  viridis,
  ggcorrplot,
  devtools,
  ragg,
  colorspace,
  sysfonts,
  showtext,
  scales
  )

options(width = 65)

knitr::opts_chunk$set(
  fig.asp = 0.618,
  fig.align = "center",
  fig.retina = 1,
  fig.showtext = T
)
```

```{r}
#| label: load-dataset-colors
#| message: false
#| echo: false
soccer <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-04-04/soccer21-22.csv')

pal_pl <- c("#F8145C", "#45F2FD", "#30FF86", "#E9FF00")
pal_pl_logo <- "#3d195b"

lab_pl_subtitle <- "Premier League match data, 2021-2022 season"

sysfonts::font_add_google("Poppins")
showtext::showtext_auto()
```

## Abstract

*Add project abstract here; delete this text when done*

## Introduction

*Add introduction here; delete this text when done*

## Question 1

What effect (if any) does being the "home" team versus being the "away" team affect the teams' performances?

<<<<<<< HEAD
### Part A
=======
## Question 1 (ii)
>>>>>>> 878466bf468d954ae346b86521e11ebcaea2555e

Do the number of goals at half-time determine the final outcome? Does this relationship change between home and away teams?

### Part B

Does home-team advantage play a major role in the final match outcome?

### Approach

Part A: The variables used to answer this question are half-time home goals (HTHG), full-time home goals (FTHG), half-time away goals (HTAG), and full-time away goals (HTHG). A scatterplot is created to quantify the correlation between goals at half-time and goals at full-time (end of match) and is color-coded based on home versus away teams. Linear regressions are performed separately for home team and away team data and correlation coefficients are reported. Scatterplot and linear regression analysis is chosen because they directly determine how strongly two variables are related to each other.

<<<<<<< HEAD
Part B: Using the Full Time Result (FTR) variable, home advantage and away advantage are categorized into a new Home_Advantage variable. The advantage is determined through the number of wins of each team. If the Result is "H", it will be added to home team victory and is therefore counted as home team advantage. If not, it will be added to away team victory and is counted as away team advantage. This information is summarized in a bar plot for Total Games to find the total goals scored by individual team in home and away. An advantage summary is created and combined together to form a new Total goals variable. The Home team, Total variable is then used as an aesthetic layer and a bar plot is created side by side to interpret the outcome for all 20 teams.

### Analysis

#### Part A: Correlation between half-time and full-time goals.
=======
*Question 1 (ii) By comparing the Full Time Result (FTR),Home Advantage, and the Away advantage combining all and summary is created with the help of Home_Advantage variable Where if the Result is "H" it will be added to home team victory otherwise it will be added to Away team victory. summarizing the output Bar plot for Total Games* *To find the total goals scored by individual team in home and away- Advantage summary is created and combined together to form a new Total goals variable. And Now with Home team, Total variable as aesthetic layers Bar plot is created side by side to interpret the outcome for all 20 teams*

\*\*

### Analysis

*Question 1 ii : As per the data_set, The home game winning advantage did not created impact in every games but only in some games whereas for the goal scored ratio in home and away match except few teams, majority of the team scored more goals in home than in away.*

#### Correlation between half-time and full-time goals.
>>>>>>> 878466bf468d954ae346b86521e11ebcaea2555e

```{r}
#| label: q1_goals
#| message: false
#| warning: false

#Create tibbles based on if the home or away team scored the goals. This makes it easier to color-code the scatterplot and calculate separate linear regressions later.
home_goals <- tibble(soccer$HTHG, soccer$FTHG)
home_goals <- home_goals %>% 
  rename(
     'half' = 'soccer$HTHG',
     'full' = 'soccer$FTHG'
  )

away_goals <- tibble(soccer$HTAG, soccer$FTAG)
away_goals <- away_goals %>% 
  rename(
    'half' = 'soccer$HTAG',
    'full' = 'soccer$FTAG'
  )

#Calculate correlation coefficients for home and away data sets.
home_r <- cor(home_goals$half, home_goals$full)
away_r <- cor(away_goals$half, away_goals$full)

#Create scatterplot with two regression lines for home and away teams, respectively. Add number of points and correlation coefficients to plot.
q1_goals <- ggplot() + 
  geom_jitter(data = home_goals, aes(x = half, y = full, color = 'Home'), width = 0.7) +
  geom_smooth(data = home_goals, aes(x = half, y = full, color = 'Home'), method = lm, se = FALSE) +
  geom_jitter(data = away_goals, aes(x = half, y = full, color = 'Away'), width = 0.7) +
  geom_smooth(data = away_goals, aes(x = half, y = full, color = 'Away'), method = lm,  se = FALSE) + 
  xlim(0, 4) +
  geom_label(aes(x = 0.25, y = 7, label= 'N = 380')) +
  geom_label(aes(x = 3.7, y = 5.5, label= 'r = 0.72', color = 'Home'), show.legend  = FALSE) +
  geom_label(aes(x = 3.7, y = 4.2, label= 'r = 0.70', color = 'Away'), show.legend = FALSE) +
  labs( x = 'Goals at half-time',
        y = 'Goals at full-time',
        color = 'Team') +
  scale_color_manual(
    breaks = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  ) + 
  theme_classic(base_size = 12) +
  theme(
    text = element_text("Poppins",
                        face = "bold",
                        color = pal_pl_logo))

#Suppress warning messages when printing ggplot object.
suppressMessages(print(q1_goals))

```

#### Part B: Correlation between half-time and full-time goals.

```{r}
#| label: q1ii_Totalgames_by_Home_and_Away
#| message: false
#| warning: false

# Create a variable for home team advantage
soccer <- soccer %>%
  mutate(HomeAdvantage = ifelse(FTR == "H", "Home Team Win", "Away Team Win"))

# Analyze home and away team advantage
advantage_summary <- soccer %>%
  group_by(HomeAdvantage) %>%
  summarise(TotalGames = n())

# Print the result
print(advantage_summary)
```

```{r}
# Bar plot for home and away team advantage

ggplot(advantage_summary, aes(x = HomeAdvantage, y = TotalGames, fill = HomeAdvantage)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Home and Away Team Wins",
       x = "Win_Count", y = "Total Games") +
  # scale_fill_manual(values = c("Home Team Win" = "lightgreen", "Away Team Win" = "red")) +
  scale_fill_manual(
    breaks = c("Home Team Win", "Away Team Win"),
    labels = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  ) +
  theme_classic() +
  theme(
    text = element_text("Poppins",
                        face = "bold",
                        color = pal_pl_logo,
                        size = 15)
  )
  # theme_minimal()
```

```{r}
total_home_goals <-soccer %>%
  group_by(HomeTeam) %>%
  summarize(Total_Home_goals = sum(FTHG))
 
total_away_goals <-soccer %>%
  group_by(AwayTeam) %>%
  summarize(Total_Away_goals = sum(FTAG))

print(total_home_goals)
print(total_away_goals)
```

```{r}
# Merge the data for home and away goals
total_goals <- merge(total_home_goals, total_away_goals, by.x = "HomeTeam", by.y = "AwayTeam", all = TRUE)

total_goals_long <- gather(total_goals, key = "TeamType", value = "TotalGoals", -HomeTeam)

# Create a side-by-side bar plot
ggplot(total_goals_long , aes(x = HomeTeam, y = TotalGoals, fill = TeamType)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(title = "Total Home and Away Goals by Team",
       x = "Teams on Home and Away", y = "Total Goals") +
  # scale_fill_manual(values = c("Total_Home_goals" = "lightgreen", "Total_Away_goals" = "red")) +
  scale_fill_manual(
    breaks = c("Total_Home_goals", "Total_Away_goals"),
    labels = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  ) +
  theme_classic() +
  theme(
    text = element_text("Poppins",
                        face = "bold",
                        color = pal_pl_logo,
                        size = 12)
  ) +
  # theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Discussion

Part A: There is a strong correlation between goals scored at half-time and goals scored at full-time. In other words, goals at half-time is a strong indicator of how many goals will be scored at the end of the match. However, there is not a perfect one-to-one relationship between these variables. For example, there are some cases where there are 0 goals scored at half-time, but 4 or 5 goals are scored by the end of the match. This plot also shows that there is slightly higher correlation between half-time and full-time goals for the home team than the away team. However, this does not necessarily determine if there is a home-team advantage because more goals does not necessarily equate to more wins.

Part B: As per the data_set, the home game winning advantage did not create impact in every game but only in some games whereas for the goal scored ratio in home and away match except few teams, majority of the team scored more goals in home than in away.

## Question 2

### Introduction

The correlation between the variables used for the analysis are first scrutinized to reveal information about them.

```{r dev = 'ragg_png'}
#| label: corr_plot
#| message: false

soccer_corr <- soccer |>
  mutate(
    HTFCR = (HR + HY)/ HF,
    ATFCR = (AR + AY/ AF)
  ) |>
  pivot_longer( cols = c( 'HF', 'AF'),
                names_to = 'Team_Fouls',
                values_to = 'Fouls') |>
  pivot_longer( cols = c( 'HY', 'AY'),
                names_to = 'Team_Yellows',
                values_to = 'yellow_cards') |>
  pivot_longer( cols = c( 'HR', 'AR'),
                names_to = 'Team_Reds',
                values_to = 'red_cards') |>
  pivot_longer(
    cols = c( 'HTFCR', 'ATFCR'),
    names_to = 'Team',
    values_to = 'FCR'
  ) |>
  pivot_longer(
    cols = c('FTHG', 'FTAG'),
    names_to = 'TFTG', values_to = 'FTG'
  )


soccer_corr <- soccer_corr |>
  select( red_cards, yellow_cards, Fouls, FCR, FTG ) |>
  rename( '🟥' = red_cards, '🟨' = yellow_cards )

cor_matrix <- round(cor(soccer_corr) , 2)

ggcorrplot(cor_matrix,
           hc.order = TRUE,
           type = 'lower',
           lab = TRUE) +
  theme_minimal() +
  theme(
    legend.position = 'none',
    panel.grid = element_blank(),
    text = element_text(
      family = "Poppins",
      face = "bold",
      color = "#3d195b"
    ),
    plot.title.position = "plot"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "The correlation between the variables used for the analysis",
    subtitle = paste0("Card to Foul Ratio(FCR), Yellow cards(🟨),\n",
                      "Red cards(🟥), and Full Time Goals (FTG)")
  )

```

*Intro goes here; delete this text when done*

### Approach

*Approach goes here; delete this text when done*

*PENALTIES BY TEAM APPROACH HERE*

To determine if there is home-team bias regarding penalties (fouls, yellow cards, and red cards), we plotted the distribution of the total number of games in the season in which a certain number of penalties were given. Although this can be shown with a density plot or area plot, it is easier to see at precisely which penalty counts were more often toward or against each team. Since most games had no red cards given at all, and the maximum number of red cards given was only two, both red cards and yellow cards are plotted on the same facet. The distribution of fouls is much greater, as every single game had at least one foul given. The two team types are mapped to both color and shape for accessibility.

### Analysis

#### Penalties by Team

```{r}
#| label: q2_penalties
#| message: false
#| fig.width: 10

q2_soccer_pen <- soccer |> 
  select(HF, AF, HY, AY, HR, AR) |>
  rename(
    yellow_home = HY,
    yellow_away = AY,
    foul_home = HF,
    foul_away = AF,
    red_home = HR,
    red_away = AR
  ) |>
  mutate(game_number = row_number()) |> 
  group_by(game_number) |> 
  pivot_longer(
    cols = c(yellow_home, red_home,
             yellow_away, red_away,
             foul_away, foul_home),
    names_to = "pen_type",
    values_to = "pen_count"
  ) |> 
  mutate(
    pen_cat = case_when(
      str_detect(pen_type, "yellow") ~ "yellow",
      str_detect(pen_type, "red") ~ "red",
      str_detect(pen_type, "foul") ~ "foul"
    ),
    h_a = case_when(
      str_detect(pen_type, "home") ~ "Home",
      str_detect(pen_type, "away") ~ "Away"
    )
  ) |>
  group_by(h_a, pen_count, pen_cat) |> 
  select(!pen_type) |> 
  summarize(tot = n()) |> 
  mutate(
    pen_cf = case_when(
      pen_cat == "foul" ~ "Fouls",
      TRUE ~ "Cards"
    )
  )

q2_penalties_viz <- ggplot(q2_soccer_pen,
                           aes(x = pen_count,
                               y = tot,
                               group = pen_cat,
                               color = h_a)) +
  facet_wrap(~pen_cf,
             ncol = 1,
             scales = "free"
             ) +
  geom_line(
    data = filter(q2_soccer_pen, pen_cat == "red"),
    aes(group = interaction(pen_cat, h_a)),
    lineend = "round",
    linewidth = 10,
    alpha = 0.05,
    color = "red"
  ) +
  geom_line(
    data = filter(q2_soccer_pen, pen_cat == "yellow"),
    aes(group = interaction(pen_cat, h_a)),
    lineend = "round",
    linewidth = 10,
    alpha = 0.1,
    color = "yellow"
  ) +
  geom_line(
    data = filter(q2_soccer_pen, pen_cat == "foul"),
    aes(group = interaction(pen_cat, h_a)),
    lineend = "round",
    linewidth = 10,
    alpha = 0.05,
    color = "purple"
  ) +
  geom_line(
    aes(group = interaction(pen_cat, h_a)),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    aes(shape = h_a),
    size = 4,
    stroke = 1.5
  ) +
  geom_text(
    data = data.frame(
      "label" = c("Red Cards", "Yellow Cards", "Fouls"),
      x = c(1, 1.75, 4),
      y = c(300, 170, 30),
      pen_cf = c("Cards", "Cards", "Fouls")
    ),
    aes(x = x, y = y, group = 1, label = label),
    color = pal_pl_logo,
    family = "Poppins",
    fontface = "bold",
    size = 5
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(7)
  ) +
  scale_color_manual(
    breaks = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  ) +
  scale_shape_manual(
    breaks = c("Home", "Away"),
    values = c(5, 4)
  ) +
  theme_classic() +
  theme(
    text = element_text("Poppins",
                        face = "bold",
                        color = pal_pl_logo,
                        size = 15),
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.85),
    legend.box.background = element_rect(),
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    plot.title.position = "plot"
  ) +
  labs(
    color = "Team",
    shape = "Team",
    x = "Total given in a game",
    y = "Number of games",
    title = "Number of English football games by amounts of penalties",
    subtitle = lab_pl_subtitle,
    fill = "Team"
  )

q2_penalties_viz
```

#### Cards:Fouls Ratio by Team

```{r}
#| label: q2_ctf
#| message: false
#| fig.width: 10

soccer_ctf <- soccer |>
  mutate(
    HTFCR = (HR + HY)/ HF,
    ATFCR = (AR + AY/ AF)
  ) |>
  rename(
    'Home' = 'HTFCR',
    'Away' = 'ATFCR'
  ) |> 
  pivot_longer(
    cols = c('Home', 'Away'),
    names_to = 'Team',
    values_to = 'FCR'
  )

ggplot(soccer_ctf, aes(x = FCR)) +
  geom_density(
    aes(fill = Team, color = Team),
    alpha = 0.25,
    linewidth = 1
  ) +
  scale_fill_manual(
    breaks = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  ) +
  scale_color_manual(
    breaks = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  ) +
  theme_classic() +
  theme(
    text = element_text("Poppins",
                        face = "bold",
                        color = pal_pl_logo,
                        size = 15),
    legend.position = "inside",
    legend.position.inside = c(0.75, 0.75),
    legend.box.background = element_rect()
  ) +
  labs(
    x = 'Card:Foul Ratio',
    y = 'Density',
    title = 'The distribution of card:foul ratios across teams',
    subtitle = lab_pl_subtitle,
    caption = 'Data from TidyTuesday'
  )
```

### Discussion

*Discussion goes here; delete this text when done*

From the chart of the number of games by amount of penalties, there does not appear to be any major trends regarding a home-team bias when it comes to Cards. That said, there are a couple more instances of the home team receiving zero cards (both yellow and red) compared to the away team. The number of fouls does vary, however. The number of instances of a team with more fouls wavers, but generally there are more instances of very high foul counts (18+) for the away team. Interestingly, there is noticeably more instances of games in which the home team received a moderate number of fouls (from 11-13). In general, it also appears that the away team is right-skewed for the total number of fouls given, while the home team is left-skewed.

The average penalties by team are provided in the following summary table:

```{r}
#| label: q2_ctf2_wrangle
#| message: false
#| echo: false

#make new Dataset
# Selecting only the necessary columns 

Compare_dataset <- soccer[, c("HomeTeam", "AwayTeam", "HF", "HR", "HY", "AF", "AR", "AY")] # Avoid division by zero by adding a small number (epsilon) to the denominator 


Compare_dataset$HCO = (Compare_dataset$HY + Compare_dataset$HR+epsilon)/Compare_dataset$HF 
# Calculating Away Foul Ratio 
Compare_dataset$ACO = (Compare_dataset$AY + Compare_dataset$AR + epsilon)/Compare_dataset$AF

# Compare_dataset

# Melt the dataset for plotting (tidyr pivot_longer is the newer alternative to gather) 

plot_data <- Compare_dataset %>% 
  pivot_longer(cols = c(HF, HR, HY, AF, AR, AY,HCO,ACO), names_to = "Category", values_to = "Count")%>%
  mutate(Type = case_when( str_detect(Category, "F") ~ "Fouls", 
                           str_detect(Category, "R") ~ "Red Cards", 
                           str_detect(Category, "Y") ~ "Yellow Cards", 
                           str_detect(Category, "C")~"Fouls Ratio"), 
         Team = case_when( str_detect(Category, "^H") ~ "Home", 
                           str_detect(Category, "^A") ~ "Away" )) %>% 
  select(-Category) %>% 
  group_by(Type, Team) %>% 
  summarise(AverageCount = mean(Count, na.rm = TRUE)) 

# Summary table
plot_data |> 
  pivot_wider(
    names_from = Type,
    values_from = AverageCount
  ) |> 
  print()
```

Based on the overall averages, the away team on average receives a greater number of penalties per game. However, it is still by a relatively slim margin. The table is shown below in chart form to visualize how close these averages are:

```{r}
#| label: q2_ctf2_viz
#| message: false
#| echo: false

ggplot(plot_data, aes(x = Type,
                      y = AverageCount,
                      fill = Team)) + 
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.9)) + 
  geom_text(
    aes(label = round(AverageCount, 2), 
        y = AverageCount + 0.02),  # Adjust the y position for visibility 
    position = position_dodge(width = 0.9), 
    vjust = 0, # Vertical adjustment; 0 means right above the bar 
    size = 5, # Text size, adjust as needed 
    color = pal_pl_logo,
    family = "Poppins",
    fontface = "bold"
  ) +
  theme_classic() +
  theme(
    text = element_text("Poppins",
                        face = "bold",
                        color = pal_pl_logo,
                        size = 15),
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.85),
    legend.box.background = element_rect(),
    plot.title.position = "plot"
  ) +
  labs(
    title = paste0("Average Fouls, Red Cards, and ",
                   "Yellow Cards\nfor Home vs. Away Teams"),
    subtitle = lab_pl_subtitle,
    x = NULL,
    y = "Count"
  ) + 
  scale_fill_manual(
    breaks = c("Home", "Away"),
    values = darken(pal_pl, 0.3)
  )
```

### 
