---
title: "Group 4 WackyWednesday - Game-Metrics: Decoding Football Outcomes Through Statistical Precision - Premier League"
subtitle: "Unveiling Team Dynamics, Discipline, and Shot Efficiency in the World of Football"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
soccer <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-04-04/soccer21-22.csv')
```

This dataset focus around one of the world's most watched sports leagues - football, It offers clear details that hold substantial potential. Football matches are known for their blend of skill, strategy, and athleticism.Its analysed through this dataset, providing a unique opportunity to explore statistical fouls affecting game outcomes. From team discipline records to the effectiveness of shots on goal, the dataset covers a wide spectrum of factors. Its richness enables a thorough examination, making it a valuable resource for gaining insights into the multifaceted elements that shape football match results and outcomes.

On the other hand, it allows for a singular exploration into the interplay between in-game dynamics and overall success or failure. This analysis could prove invaluable to observers, coaches, and even players striving to enhance their tactical approaches.

By incorporating halftime data, the analysis gains additional depth, providing a fresh perspective on the dynamics within games. In Conclusion, this dataset offers a comprehensive view of the multifaceted elements contributing to the outcome of football matches. Its richness and depth make it a valuable resource for anyone seeking to gain insights into the complexities of the game, be it for strategic planning, performance improvement, or a deeper understanding of the sport's intricacies.

Make sure to load the data and use inline code for some of this information.

Text HERE.

```{r}
#| label: explore-dataset
## Exploratory code HERE.
```

## Questions

The two questions you want to answer.

### Question 1

The first topic we will explore is if being on the home or away team has a significant effect on the team's performance. Our hypothesis is that the home team will have 1) more goals at both half-time and full-time and 2) more wins due to "home team advantage", where it is thought that the hosting team has a significant advantage over the visiting team. This hypothesis will be explored and tested using the variables described in "Analysis plan".

### Question 2

Text HERE.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

### Question 1

-   Variables involved:
    -   HERE
-   Variables to be created:
    -   HERE (if applicable)
-   External data:
    -   HERE (if applicable)

### Question 2

-   Variables involved:
    -   Goals: `FTHG` (Full Time Home Goals), `FTAG` (Full Time Away Goals), and `FTR` (Full Time Results).
    -   Fouls: `HF` (Home Fouls) and `AF` (Away Fouls).
    -   Cards: `HY` (Home Yellow) , `AY` (Away Yellow), `HR` (Home Red), and `AR` (Away Red).
-   Variables to be created:
    -   Ratios between the total number of yellow cards, red cards, and fouls, by team and game.
    -   Ratio of total number of cards and number of fouls, by team and game.
-   This question is not likely to require external data to be merged in.
