---
title: "Premier League - Group 4 WackyWednesday"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
soccer <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-04-04/soccer21-22.csv')
```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

Text HERE.

```{r}
#| label: explore-dataset
## Exploratory code HERE.
```

## Questions

The two questions you want to answer.

### Question 1

The first topic we will explore is if being on the home or away team has a significant effect on the team's performance. Our hypothesis is that the home team will have 1) more goals at both half-time and full-time and 2) more wins due to "home team advantage", where it is thought that the hosting team has a significant advantage over the visiting team. This hypothesis will be explored and tested using the variables described in "Analysis plan".

### Question 2

Text HERE.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

### Question 1

-   Variables involved:
    -   HERE
-   Variables to be created:
    -   HERE (if applicable)
-   External data:
    -   HERE (if applicable)

### Question 2

-   Variables involved:
    -   Goals: `FTHG` (Full Time Home Goals), `FTAG` (Full Time Away Goals), and `FTR` (Full Time Results).
    -   Fouls: `HF` (Home Fouls) and `AF` (Away Fouls).
    -   Cards: `HY` (Home Yellow) , `AY` (Away Yellow), `HR` (Home Red), and `AR` (Away Red).
-   Variables to be created:
    -   Ratios between the total number of yellow cards, red cards, and fouls, by team and game.
    -   Ratio of total number of cards and number of fouls, by team and game.
-   This question is not likely to require external data to be merged in.
